Multi-level generalizations of collective spin dynamics and squeezing

### 2019-10-17 collective_sun/multilevel_spin_notes.tex
capitalized all single-body spin operators, added expression of the interaction Hamiltonian in terms of permutation operators

\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\T}{\mathcal{T}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\X}{\mathcal{X}}

@@

@@

@@

We consider a collection of $n$-level fermions on a 1-D lattice, with
a single-particle Hamiltonian
\begin{align}
  H_{\t{lat}} = -J \sum_{q,\mu} \cos\p{qa} c_{q\mu}^\dag c_{q\mu},
  \label{eq:H_lat}
\end{align}
where $J$ is the nearest-neighbor tunneling rate, $a$ is the spacing
between neighboring lattice sites, and $c_{q\mu}$ is the fermionic
annihilation operator for a particle in quasi-momentum mode $q$ with
internal state (e.g.~nuclear spin) $\mu$.  Under the frozen-mode
approximation, collective two-body SU($n$)-symmetric interactions
between these fermions take form\cite{perlin2019effective}
\begin{align}
  H_{\t{int}} = \f{u}{2} \sum_{p,q,\mu,\nu}
  \p{c_{p\mu}^\dag c_{q\nu}^\dag c_{q\nu} c_{p\mu}
    + c_{q\mu}^\dag c_{p\nu}^\dag c_{q\nu} c_{p\mu}},
\end{align}
where $u\equiv U/L$ is the two-body on-site interaction energy $U$
divided by the number of lattice sites $L$.  Defining the spin
operators $s_{\mu\nu}^{(p)}\equiv c_{p\mu}^\dag c_{p\nu}$ for each
quasi-momentum $p$, we can alternately write
\begin{align}
  H_{\t{int}} = \f{u}{2} \sum_{p,q,\mu,\nu}
  \p{s_{\mu\mu}^{(p)} s_{\nu\nu}^{(q)} - s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)}}
  = \f{u}{2}\p{N^2 - \sum_{p,q,\mu,\nu} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)}},
\end{align}
where $N$ is the total number of particles.  In the absence of
coherence between sectors of different particle number, we can neglect
the $\sim N^2$ term and simply write
\begin{align}
  H_{\t{int}}
  = - \f{u}{2} \sum_{p,q,\mu,\nu} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)}.
\end{align}
In the case of SU(2), at this point we would expand the spin operators
$s_{\mu\nu}^{(p)}$ in terms of Pauli operators in order to convert
$H_{\t{int}}$ into an SU(2) spin Hamiltonian\cite{he2019engineering}.
To generalize this procedure to the case of SU($n$), we first define a
vector $\v s^{(p)}$ of all spin operators $s_{\mu\nu}^{(p)}$, and
write
\begin{align}
  H_{\t{int}}
  = - \f{u}{2} \sum_{p,q} {\v s^{(p)}}^\dag \v s^{(q)}
  = - \f{u}{2} \v\S^\dag \c \v\S,
  &&
  \v\S\equiv \sum_p \v s^{(p)}.
  \label{eq:H_int_spin}
\end{align}
In the remainder of this section, we identify various bases for the
space of linear operators on the Hilbert space $\H_n$ of an $n$-level
system.  For brevity, we denote this space of operators by
$\B\p{\H_n}$, and note that $\B\p{\H_n}$ is itself an
$n^2$-dimensional Hilbert space equipped with a trace
(Hilbert-Schmidt) inner product
\begin{align}
  \obk{\O|\Q} \equiv \tr\p{\O^\dag Q},
\end{align}
where $\oket{\Q}$ and $\obra{\O}$ respectively denote vectors in
$\B\p{\H_n}$ and its dual space $\B\p{\H_n}^*$.  Similarly to the set
of all spin operators $s_{\mu\nu}\equiv\op{\mu}{\nu}$ on the
single-particle Hilbert space at each quasi-momentum, all of the bases
we consider will be orthonormal with respect to the trace inner
product.  As we show in Appendix \ref{sec:changing_bases}, the
collective SU($n$)-symmetric interaction Hamiltonian in
\eqref{eq:H_int_spin} takes an identical form in any orthonormal basis
for $\B\p{\H_n}$.

@@

The most obvious generalization of Pauli operators to the case of
SU($n$) are the {\it generalized Gell-Mann (GGM)
  operators}\cite{hioe1981level, bertlmann2008bloch}, defined for
$j,k,\ell\in\set{0,1,\cdots,n-1}$ with $k<j$ and $\ell>0$ by
\begin{align}
  \lambda_0 \equiv \sqrt{\f{2}{n}}~ \1,
  &&
  \lambda_\ell \equiv \sqrt{\f{2}{\ell\p{\ell+1}}}
  \p{\sum_{n=0}^{\ell-1}\op{n} - \ell \op{\ell}},
  \label{eq:GGM_diag}
\end{align}
\begin{align}
  \lambda_{jk,+} \equiv \op{j}{k} + \op{k}{j},
  &&
  \lambda_{jk,-} \equiv i\p{\op{j}{k} - \op{k}{j}},
  \label{eq:GGM_off_diag}
\end{align}
where $\1$ is the identity operator.  The self-adjoint GGM operators
in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag} provide a complete
basis for the space of operators on the Hilbert space $\H_n$ of an
$n$-level system, coincide exactly with the Pauli operators in the
case of $n=2$, and satisfy the orthonormality condition
\begin{align}
  \obk{\lambda_a|\lambda_b} = 2\delta_{ab},
  \label{eq:GGM_inner}
\end{align}
where $a,b$ index any GGM operator in \eqref{eq:GGM_diag} or
\eqref{eq:GGM_off_diag}.  Denoting a vector of all GGM operators by
$\v\lambda$, we can therefore write (see Appendix
\ref{sec:changing_bases})
\begin{align}
  H_{\t{int}}
  = -\f{u}{4} \sum_{p,q} \v\lambda^{(p)} \c \v\lambda^{(q)}
  = -\f{u}{4} \v\Lambda \c \v\Lambda,
  &&
  \v\Lambda \equiv \sum_p \v\lambda^{(p)}.
  \label{eq:H_int_GGM}
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Classifying collective states}
\label{sec:classifying}

Given a quantum system composed of $N$ identical (but distinguishable)
subsystems, e.g.~an array of $n$-level nuclear spins, we identify the
set $\Pi\equiv\set{\Pi_\sigma}$ of operators that permute subsystems.
Permutation operators $\Pi_\sigma$ are labeled by elements
$\sigma\in\SS_N$ of the symmetric group $\SS_N$ of order $N$.  We
refer to simultaneous eigenstates of all $\Pi_\sigma\in\Pi$ as {\it
  collective states}.  The subset of collective states that have
eigenvalue 1 with respect to all permutation operators
$\Pi_\sigma\in\Pi$ is the {\it maximally symmetric collective (MSC)}
manifold, which is precisely the Dicke manifold in the case of
two-level subsystems ($n=2$).  The MSC manifold is spanned by a basis
$\set{\ket{\v m}}$ of MSC states that can be uniquely identified by a
list of integers $\v m=\p{m_0,m_1,\cdots,m_{n-1}}\in\ZZ_{N+1}^n$,
where the integer $m_j\in\ZZ_{N+1}$ ($0\le m_j\le N$) indicates the
total occupation number of subsystem state $j$, and $\sum_j m_j=N$.
The Dicke manifold of $N$ qubits, for example, is spanned by the Dicke
states $\set{\ket{m_0,m_1}}$, where $m_0+m_1=N$ and the integer $m_0$
($m_1$) indicates the occupation number of the qubit state $\ket{0}$
($\ket{1}$).

The MSC manifold has two important properties that concern us.  First,
the MSC manifold is closed under the action of unitaries generated by
linear combinations collective drive operators $\D_{LM}$.  Second, the
MSC manifold is precisely the ground-state manifold of the collective
SU($n$)-symmetric interaction Hamiltonian $H_{\t{int}}$ in
\eqref{eq:H_int_spin} or \eqref{eq:H_int_drive}.  We prove these
properties of the MSC manifold in Appendix \ref{sec:MSC_proofs}.

@@

The collective interaction Hamiltonian in \eqref{eq:H_int_spin} takes
the form
\begin{align}
  H = \sum_{p,q,j} X_j^{(p)} {X_j^{(q)}}^\dag,
\end{align}
which essentially consists of two-body interactions ($p\ne q$) and
single-body terms ($p=q$), respectively of the form
\begin{align}
  \sum_j X_j \otimes X_j^\dag,
  &&
  \sum_j X_j X_j^\dag.
\end{align}
Choosing a different basis $Y\equiv\set{Y_j}$ for $\B\p{\H_n}$ with
corresponding norm $\N_Y$, we can expand
\begin{align}
  X_j = \f1{\N_Y} \sum_k Y_k \obk{Y_k|X_j},
  &&
  X_j^\dag = \f1{\N_Y} \sum_k Y_k^\dag \obk{Y_k^\dag|X_j^\dag}
  = \f1{\N_Y} \sum_k Y_k^\dag \obk{X_j|Y_k},
\end{align}
where we used the fact that
\begin{align}
  \obk{Y_k^\dag|X_j^\dag}
  = \tr\p{Y_k X_j^\dag}
  = \tr\p{X_j^\dag Y_k}
  = \obk{X_j|Y_k}.
\end{align}
By resolution of the identity, we can therefore find
\begin{align}
  \sum_j X_j \otimes X_j^\dag
  = \f1{\N_Y^2} \sum_{j,k,\ell} Y_k \obk{Y_k|X_j}
  \otimes Y_\ell^\dag \obk{X_j|Y_\ell}
  = \f{\N_X}{\N_Y^2} \sum_{k,\ell} Y_k \otimes Y_\ell^\dag
  \obk{Y_k|\I|Y_\ell}
  = \f{\N_X}{\N_Y} \sum_k Y_k \otimes Y_k^\dag,
\end{align}
and similarly
\begin{align}
  \sum_j X_j X_j^\dag
  = \f1{\N_Y^2} \sum_{j,k,\ell} Y_k Y_\ell^\dag
  \obk{Y_k|X_j} \obk{X_j|Y_\ell}
  = \f{\N_X}{\N_Y} \sum_k Y_k Y_k^\dag.
\end{align}
The collective interaction Hamiltonian in \eqref{eq:H_int_spin} thus
takes an identical form in any orthonormal basis for the space
$\B\p{\H_n}$ of linear operators on the Hilbert space $\H_n$ of an
$n$-level system, up to appropriate re-scaling factors $\N_X/\N_Y$.

@@

Here we derive the drive-operator ($D_{LM}$) expansion of the
spin-orbit coupled Hamiltonian $H_{\t{lat}}^{(\phi)}$ in
\eqref{eq:H_lat_SOC}.  In the basis of spin operators
$s_{\mu\mu}^{(q)} \equiv c_{q\mu}^\dag c_{q\mu}$, this Hamiltonian
takes the form
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} s_{\mu\mu}^{(q)}
  \label{eq:H_lat_spin}
\end{align}
Defining the diagonal ``spin-wave'' operators
\begin{align}
  \tilde d_{\phi+} \equiv \sum_\mu \cos\p{\mu\phi} s_{\mu\mu},
  &&
  \tilde d_{\phi-} \equiv \sum_\mu \sin\p{\mu\phi} s_{\mu\mu},
\end{align}
which are respectively even and odd under spin inversion
($\mu\to-\mu$), we can expand
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \tilde d_{\phi+} - \sin\p{qa} \tilde d_{\phi-},
\end{align}
and resolve the identity (super-)operator $\I$ with
$\I \tilde d_{\phi\pm} = \tilde d_{\phi\pm}$ in the basis of the drive operators
$D_{LM}$ (see Appendix \ref{sec:changing_bases}), finding
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L,M} \obk{\tilde d_{\phi+}|D_{LM}} D_{LM}
  - \sin\p{qa} \f12 \sum_{L,M} \obk{\tilde d_{\phi-}|D_{LM}} D_{LM}.
\end{align}
We now note that the drive operators $D_{LM}$ are strictly
off-diagonal for $M\ne0$, and furthermore that $D_{L,0}$ with even
(odd) $L$ is even (odd) under spin inversion, which implies
\begin{align}
  \sum_{L,M} \obk{\tilde d_{\phi+}|D_{LM}} D_{LM}
  = \sum_{L\,\t{even}} \obk{\tilde d_{\phi+}|D_{L,0}} D_{L,0},
\end{align}
and likewise with $\tilde d_{\phi-}$ and odd $L$.  In total, we can
write
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L\,\t{even}}
  \obk{\tilde d_{\phi+}|D_{L,0}} D_{L,0}
  - \sin\p{qa} \f12 \sum_{L\,\t{odd}}
  \obk{\tilde d_{\phi-}|D_{L,0}} D_{L,0},
\end{align}
or, more compactly,
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} s_{\mu\mu}
  = \sum_L w_L\p{qa} A_L^{(\phi)} D_{L,0},
  &&
  w_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases},
  \label{eq:compact_SOC_drive}
\end{align}
with effective driving field amplitudes
\begin{align}
  A_L^{(\phi)}
  \equiv \p{-1}^L \f12 \obk{\tilde d_{\phi,\p{-1}^L}|D_{L,0}}
  = \p{-1}^L \sqrt{\f{2L+1}{2I+1}}
  \sum_\mu \bk{I\mu;L,0|I\mu} w_L\p{\mu\phi}.
\end{align}
Substituting the result in \eqref{eq:compact_SOC_drive} into the
single-particle Hamiltonian in \eqref{eq:H_lat_spin} yields
\begin{align}
  H_{\t{lat}}^{(\phi)} = \sum_{q,L} B_L^{(\phi,q)} D_{L,0}^{(q)},
  &&
  B_L^{(\phi,q)} \equiv -J w_L\p{qa} A_L^{(\phi)}.
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Key properties of the maximally-symmetric collective
  manifold}
\label{sec:MSC_proofs}

Here we prove two important properties of the maximally-symmetric
collective (MSC) manifold defined in Section \ref{sec:classifying}:
\begin{enumerate*}
\item \label{pt:MSC_closure} that the MSC manifold is closed under the
  action of unitaries generated by linear combinations collective
  drive operators $\D_{LM}$; and
\item \label{pt:MSC_ground} that the MSC manifold is precisely the
  ground-state manifold of the collective SU($n$)-symmetric
  interaction Hamiltonian $H_{\t{int}}$ in \eqref{eq:H_int_spin} or
  \eqref{eq:H_int_drive}.
\end{enumerate*}
Our starting ingredients are an $n$-dimensional Hilbert space $\H_n$,
its $N$-fold tensor product $\H_n^{\otimes N}$, and the set
$\Pi\equiv\set{\Pi_\sigma}$ of operators that permute the tensor
factors of $\H_n^{\otimes N}$.  Permutation operators $\Pi_\sigma$ are
labeled by elements $\sigma\in\SS_N$ of the symmetric group $\SS_N$ of
order $N$.  The MSC sub-manifold of $\H_n^{\otimes N}$ is defined by
the set of states that have eigenvalue $1$ with respect to all
permutation operators $\Pi_\sigma\in\Pi$.

Given an orthonormal basis $X=\set{X_j}$ for the space $\B\p{\H_n}$ of
linear operators on $\H_n$, we define the collective operator
$\X_j\in\B\p{\H_n^{\otimes N}}$ by
\begin{align}
  \X_j \equiv \sum_{p\in\ZZ_N} X_j^{(p)},
\end{align}
where $\ZZ_N\equiv\set{0,1,\cdots,N-1}$ and $X_j^{(p)}$ acts with
$X_j$ on tensor factor $p$ of $\H_n^{\otimes N}$ and trivially on all
other tensor factors, i.e.
\begin{align}
  X_j^{(p)} \equiv
  \1^{\otimes p} \otimes X_j \otimes \1^{\otimes\p{N-p-1}}.
\end{align}
We denote a vector of all collective operators in $\X\equiv\set{\X_j}$
by $\v\X$, i.e.~$\v\X\equiv\p{\X_j:\X_j\in\X}$.

>>>>

\newcommand{\B}{\mathcal{B}}
\newcommand{\D}{\mathcal{D}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\T}{\mathcal{T}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\X}{\mathcal{X}}

@@

\newcommand{\MSC}{\text{MSC}}

@@

%%% text markup
\usepackage{color} % text color
\newcommand{\red}[1]{{\color{red} #1}}

@@

We consider a collection of $n$-level fermions on a 1-D lattice, with
a single-particle Hamiltonian
\begin{align}
  H_{\t{lat}} = -J \sum_{q,\mu} \cos\p{qa} c_{q\mu}^\dag c_{q\mu},
  \label{eq:H_lat}
\end{align}
where $J$ is the nearest-neighbor tunneling rate, $a$ is the spacing
between neighboring lattice sites, and $c_{q\mu}$ is the fermionic
annihilation operator for a particle in quasi-momentum mode $q$ with
internal state (e.g.~nuclear spin) $\mu$.  Under the frozen-mode
approximation, collective two-body SU($n$)-symmetric interactions
between these fermions take form\cite{perlin2019effective}
\begin{align}
  H_{\t{int}} = \f{u}{2} \sum_{p,q,\mu,\nu}
  \p{c_{p\mu}^\dag c_{q\nu}^\dag c_{q\nu} c_{p\mu}
    + c_{q\mu}^\dag c_{p\nu}^\dag c_{q\nu} c_{p\mu}},
\end{align}
where $u\equiv U/L$ is the two-body on-site interaction energy $U$
divided by the number of lattice sites $L$.  Defining the spin
operators $S_{\mu\nu}^{(p)}\equiv c_{p\mu}^\dag c_{p\nu}$ for each
quasi-momentum $p$, we can alternately write
\begin{align}
  H_{\t{int}} = \f{u}{2} \sum_{p,q,\mu,\nu}
  \p{S_{\mu\mu}^{(p)} S_{\nu\nu}^{(q)} - S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}}
  = \f{u}{2}\p{N^2 - \sum_{p,q,\mu,\nu}
    S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}},
\end{align}
where $N$ is the total number of particles.  In the absence of
coherence between sectors of different particle number, we can neglect
the $\sim N^2$ term and simply write
\begin{align}
  H_{\t{int}}
  = - \f{u}{2} \sum_{p,q,\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}.
\end{align}
In the case of SU(2), at this point we would expand the spin operators
$S_{\mu\nu}^{(p)}$ in terms of Pauli operators in order to convert
$H_{\t{int}}$ into an SU(2) spin Hamiltonian\cite{he2019engineering}.
To generalize this procedure to the case of SU($n$), we first define a
vector $\v S^{(p)}$ of all spin operators $S_{\mu\nu}^{(p)}$, and
write
\begin{align}
  H_{\t{int}}
  = - \f{u}{2} \sum_{p,q} {\v S^{(p)}}^\dag \c \v S^{(q)}
  = - \f{u}{2} \v\S^\dag \c \v\S,
  &&
  \v\S\equiv \sum_p \v S^{(p)}.
  \label{eq:H_int_spin}
\end{align}
In the remainder of this section, we identify various bases for the
space of linear operators on the Hilbert space $\H_n$ of an $n$-level
system.  For brevity, we denote this space of operators by
$\B\p{\H_n}$, and note that $\B\p{\H_n}$ is itself an
$n^2$-dimensional Hilbert space equipped with a trace
(Hilbert-Schmidt) inner product
\begin{align}
  \obk{\O|\Q} \equiv \tr\p{\O^\dag Q},
\end{align}
where $\oket{\Q}$ and $\obra{\O}$ respectively denote vectors in
$\B\p{\H_n}$ and its dual space $\B\p{\H_n}^*$.  Similarly to the set
of all spin operators $S_{\mu\nu}\equiv\op{\mu}{\nu}$ on the
single-particle Hilbert space at each quasi-momentum, all of the bases
we consider will be orthonormal with respect to the trace inner
product.  As we show in Appendix \ref{sec:changing_bases}, the
collective SU($n$)-symmetric interaction Hamiltonian in
\eqref{eq:H_int_spin} takes an identical form in any orthonormal basis
for $\B\p{\H_n}$.

@@

The most obvious generalization of Pauli operators to the case of
SU($n$) are the {\it generalized Gell-Mann (GGM)
  operators}\cite{hioe1981level, bertlmann2008bloch}, defined for
$j,k,\ell\in\set{0,1,\cdots,n-1}$ with $j<k$ and $\ell>0$ by
\begin{align}
  \lambda_0 \equiv \sqrt{\f{2}{n}}~ \1,
  &&
  \lambda_\ell \equiv \sqrt{\f{2}{\ell\p{\ell+1}}}
  \p{\sum_{n=0}^{\ell-1}\op{n} - \ell \op{\ell}},
  \label{eq:GGM_diag}
\end{align}
\begin{align}
  \lambda_{jk,\x} \equiv \op{j}{k} + \op{k}{j},
  &&
  \lambda_{jk,\y} \equiv i\p{\op{j}{k} - \op{k}{j}},
  \label{eq:GGM_off_diag}
\end{align}
where $\1$ is the identity operator.  The self-adjoint GGM operators
in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag} provide a complete
basis for the space of operators on the Hilbert space $\H_n$ of an
$n$-level system, coincide exactly with the Pauli operators in the
case of $n=2$, and satisfy the orthonormality condition
\begin{align}
  \obk{\lambda_a|\lambda_b} = 2\delta_{ab},
  \label{eq:GGM_inner}
\end{align}
where $a,b$ index any GGM operator in \eqref{eq:GGM_diag} or
\eqref{eq:GGM_off_diag}.  Denoting a vector of all GGM operators by
$\v\lambda$, we can therefore write (see Appendix
\ref{sec:changing_bases})
\begin{align}
  H_{\t{int}}
  = -\f{u}{4} \sum_{p,q} \v\lambda^{(p)} \c \v\lambda^{(q)}
  = -\f{u}{4} \v\Lambda \c \v\Lambda,
  &&
  \v\Lambda \equiv \sum_p \v\lambda^{(p)}.
  \label{eq:H_int_GGM}
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Collective states}
\label{sec:collective}

Given a quantum system composed of $N$ identical (but distinguishable)
$n$-level subsystems, e.g.~an array of $n$-level nuclear spins, we
identify the set $\Pi\equiv\set{\Pi_\sigma}$ of operators that permute
subsystems, labeled by corresponding elements $\sigma\in\SS_N$ of the
symmetric group $\SS_N$ of order $N$.  We refer to simultaneous
eigenstates of all $\Pi_\sigma\in\Pi$ as {\it collective states}.  The
subset of collective states that have eigenvalue $1$ with respect to
all permutation operators $\Pi_\sigma\in\Pi$ is the {\it maximally
  symmetric collective (MSC)} manifold $\M_\MSC$, which is precisely
the Dicke manifold in the case of two-level subsystems ($n=2$).  The
MSC manifold is spanned by a basis $\set{\ket{\v m}}$ of MSC states
that can be uniquely identified by a list of integers
$\v m=\p{m_0,m_1,\cdots,m_{n-1}}\in\ZZ_{N+1}^n$, where $m_k$ indicates
the total occupation number of subsystem state $k$, and
$\sum_k m_k=N$.  The Dicke manifold of $N$ qubits, for example, is
spanned by the Dicke states $\set{\ket{m_0,m_1}}$, where $m_0+m_1=N$
and the integer $m_0$ ($m_1$) indicates the occupation number of the
qubit state $\ket{0}$ ($\ket{1}$).  The dimension of the MSC manifold
is determined by the number of ways to assign $N$ sub-systems to $n$
states, which is ${N+n-1 \choose n-1}$.

The MSC manifold has two key properties.  First, the MSC manifold is
closed under the action of unitaries generated by collective drive
operators $\D_{LM}$, as well as linear combinations and products
thereof.  Closure of the MSC manifold under collective dynamics is a
straightforward consequence of the permutational symmetry obeyed by
collective operators.  Second, the MSC manifold is the ground-state
manifold of the collective SU($n$)-symmetric interaction Hamiltonian
$H_{\t{int}}$ in \eqref{eq:H_int_spin}.  To illuminate the spectral
properties of $H_{\t{int}}$, we can expand
\begin{align}
  H_{\t{int}}
  = -\f{u}{2} \sum_{p,q,\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}
  = -u\sum_{\substack{p<q\\\mu,\nu}} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(q)}
  - \f{u}{2} \sum_{p,\mu,\nu} S_{\mu\nu}^{(p)} S_{\nu\mu}^{(p)}
  = -u\sum_{p<q} \Pi_{pq} - \f12 N n u
  \simeq -u\sum_{p<q} \Pi_{pq},
  \label{eq:H_int_perm}
\end{align}
where $\Pi_{pq}\in\Pi$ permutes subsystems $p$ and $q$, and $\simeq$
denotes equality up to an overall constant with no physical
consequence.  The form of $H_{\t{int}}$ in \eqref{eq:H_int_perm}
reveals that its eigenstates can be classified by their symmetries
under the permutation of subsystems.  In particular, the ground-state
manifold of $H_{\t{int}}$ consists of states that have eigenvalue 1
with respect to all pair-wise permutation operators $\Pi_{pq}$, which
is precisely the MSC manifold $\M_\MSC$.  The interaction energy of
states $\ket{\psi_\MSC}\in\M_\MSC$ is (up to an overall constant)
simply $-{ N\choose 2 } u = -\p{1/4}N\p{N-1} u$.

@@

The collective interaction Hamiltonian in \eqref{eq:H_int_spin} takes
the form
\begin{align}
  H = \sum_{p,q,j} X_j^{(p)} {X_j^{(q)}}^\dag,
\end{align}
which essentially consists of single-body terms ($p=q$) and two-body
interactions ($p\ne q$) of the form
\begin{align}
  \sum_j X_j X_j^\dag,
  &&
  \sum_j X_j \otimes X_j^\dag.
\end{align}
Choosing an arbitrary orthonormal basis $Y\equiv\set{Y_j}$ for
$\B\p{\H_n}$ with corresponding norm $\N_Y$, we can expand
\begin{align}
  X_j = \f1{\N_Y} \sum_k Y_k \obk{Y_k|X_j},
  &&
  X_j^\dag = \f1{\N_Y} \sum_k Y_k^\dag \obk{Y_k^\dag|X_j^\dag}
  = \f1{\N_Y} \sum_k Y_k^\dag \obk{X_j|Y_k},
\end{align}
where we used the fact that
\begin{align}
  \obk{Y_k^\dag|X_j^\dag}
  = \tr\p{Y_k X_j^\dag}
  = \tr\p{X_j^\dag Y_k}
  = \obk{X_j|Y_k}.
\end{align}
Denoting an arbitrary bilinear operation by $\odot$, by resolution of
the identity we therefore find
\begin{align}
  \sum_j X_j \odot X_j^\dag
  = \f1{\N_Y^2} \sum_{j,k,\ell} Y_k \obk{Y_k|X_j}
  \odot Y_\ell^\dag \obk{X_j|Y_\ell}
  = \f{\N_X}{\N_Y^2} \sum_{k,\ell} Y_k \odot Y_\ell^\dag
  \obk{Y_k|\I|Y_\ell}
  = \f{\N_X}{\N_Y} \sum_k Y_k \odot Y_k^\dag,
\end{align}
which in particular implies that
\begin{align}
  \sum_j X_j X_j^\dag
  = \f{\N_X}{\N_Y} \sum_k Y_k Y_k^\dag,
  &&
  \sum_j X_j \otimes X_j^\dag
  = \f{\N_X}{\N_Y} \sum_k Y_k \otimes Y_k^\dag,
\end{align}
and in turn
\begin{align}
  H = \f{\N_X}{\N_Y} \sum_{p,q,j} Y_j^{(p)} {Y_j^{(q)}}^\dag.
\end{align}
The collective interaction Hamiltonian in \eqref{eq:H_int_spin} thus
takes an identical form in any orthonormal basis for the space
$\B\p{\H_n}$ of linear operators on the Hilbert space $\H_n$ of an
$n$-level system, up to appropriate re-scaling factors $\N_X/\N_Y$.

@@

Here we derive the drive-operator ($D_{LM}$) expansion of the
spin-orbit coupled Hamiltonian $H_{\t{lat}}^{(\phi)}$ in
\eqref{eq:H_lat_SOC}.  In the basis of spin operators
$S_{\mu\mu}^{(q)} \equiv c_{q\mu}^\dag c_{q\mu}$, this Hamiltonian
takes the form
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} S_{\mu\mu}^{(q)}
  \label{eq:H_lat_spin}
\end{align}
Defining the diagonal ``spin-wave'' operators
\begin{align}
  \tilde d_{\phi+} \equiv \sum_\mu \cos\p{\mu\phi} S_{\mu\mu},
  &&
  \tilde d_{\phi-} \equiv \sum_\mu \sin\p{\mu\phi} S_{\mu\mu},
\end{align}
which are respectively even and odd under spin inversion
($\mu\to-\mu$), we can expand
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} S_{\mu\mu}
  = \cos\p{qa} \tilde d_{\phi+} - \sin\p{qa} \tilde d_{\phi-},
\end{align}
and resolve the identity (super-)operator $\I$ with
$\I \tilde d_{\phi\pm} = \tilde d_{\phi\pm}$ in the basis of the drive operators
$D_{LM}$ (see Appendix \ref{sec:changing_bases}), finding
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} S_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L,M} \obk{\tilde d_{\phi+}|D_{LM}} D_{LM}
  - \sin\p{qa} \f12 \sum_{L,M} \obk{\tilde d_{\phi-}|D_{LM}} D_{LM}.
\end{align}
We now note that the drive operators $D_{LM}$ are strictly
off-diagonal for $M\ne0$, and furthermore that $D_{L,0}$ with even
(odd) $L$ is even (odd) under spin inversion, which implies
\begin{align}
  \sum_{L,M} \obk{\tilde d_{\phi+}|D_{LM}} D_{LM}
  = \sum_{L\,\t{even}} \obk{\tilde d_{\phi+}|D_{L,0}} D_{L,0},
\end{align}
and likewise with $\tilde d_{\phi-}$ and odd $L$.  In total, we can
write
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} S_{\mu\mu}
  = \cos\p{qa} \f12 \sum_{L\,\t{even}}
  \obk{\tilde d_{\phi+}|D_{L,0}} D_{L,0}
  - \sin\p{qa} \f12 \sum_{L\,\t{odd}}
  \obk{\tilde d_{\phi-}|D_{L,0}} D_{L,0},
\end{align}
or, more compactly,
\begin{align}
  \sum_\mu \cos\p{qa+\mu\phi} S_{\mu\mu}
  = \sum_L w_L\p{qa} A_L^{(\phi)} D_{L,0},
  &&
  w_L\p{\theta} \equiv
  \begin{cases}
    \cos\theta & L~\t{even} \\
    \sin\theta & L~\t{odd}
  \end{cases},
  \label{eq:compact_SOC_drive}
\end{align}
with effective driving field amplitudes
\begin{align}
  A_L^{(\phi)}
  \equiv \p{-1}^L \f12 \obk{\tilde d_{\phi,\p{-1}^L}|D_{L,0}}
  = \p{-1}^L \sqrt{\f{2L+1}{2I+1}}
  \sum_\mu \bk{I\mu;L,0|I\mu} w_L\p{\mu\phi}.
\end{align}
Substituting the result in \eqref{eq:compact_SOC_drive} into the
single-particle Hamiltonian in \eqref{eq:H_lat_spin} yields
\begin{align}
  H_{\t{lat}}^{(\phi)} = \sum_{q,L} B_L^{(\phi,q)} D_{L,0}^{(q)},
  &&
  B_L^{(\phi,q)} \equiv -J w_L\p{qa} A_L^{(\phi)}.
\end{align}

@@
